Sample variance in weak lensing: how many simulations are required?

### 2016-03-02 Paper/paper.tex
incorporated comments from referee

\begin{equation}
\label{covest}
\bbh{C} = \frac{1}{N_r-1}\sum_{r=1}^{N_r} (\bbh{d}_r - \bar{\bb{d}}) (\bbh{d}_r - \bar{\bb{d}})^T.
\end{equation}
%
Assuming a normal feature likelihood, together with a flat prior on
the parameter space, the parameter posterior distribution
$\mathcal{L}(\bb{p}\vert\bbh{d}_{\rm obs})$ given an observed instance of
$\bbh{d}$, which we call $\bbh{d}_{\rm obs}$, follows from Bayes' theorem,
%
\begin{equation}
\label{posteriorbayes}
-2\log\mathcal{L}(\bb{p}\vert\bbh{d}_{\rm obs}) = [\bbh{d}_{\rm obs}-\bb{d}(\bb{p})]^T\bbh{C}^{-1}[\bbh{d}_{\rm obs}-\bb{d}(\bb{p})].
\end{equation}
%
For the sake of simplicity, we approximate the posterior as a Gaussian
around its maximum. This corresponds to Taylor-expanding the simulated
feature to first order around a point $\bb{p}_0$ (ideally the maximum
of Eq.~\ref{posteriorbayes}):
%
\begin{equation}
\bb{d}(\bb{p}) \approx \bb{d}_0 + \bb{d}_0^\prime(\bb{p}-\bb{p}_0).
\end{equation}
%

To measure the derivatives of the features $\bb{d}'_0$ with respect to
the cosmological parameters, we make use of the public code
\texttt{Nicaea} \citep{Nicaea} for the power spectrum, and we use an
independent simulation set (containing simulations with a variety of
different combinations of $(\Omega_m,w,\sigma_8)$, see Table
\ref{simtable}) for the peak counts.

@@

\begin{equation}
\label{estimatorfisher}
\bbh{F} = \bb{d}_0'^T\bbh{\Psi}\bb{d}_0'.
\end{equation}
%
Here we introduced the familiar Fisher matrix estimator $\bbh{F} =
\bb{d}_0'^T\bbh{\Psi}\bb{d}_0'$ and, for simplicity, we assumed
$\langle\bbh{d}_{\rm obs}\rangle_O=\bb{d}_0$, so that
$\langle(\bbh{d}_{\rm obs}-\bb{d}_0)(\bbh{d}_{\rm obs}-\bb{d}_0)^T\rangle_O=\bb{C}$.  When we perform an observation
$\bbh{d}_{\rm obs}$, the parameter estimate $\bbh{p}$ is a random draw
from a probability distribution with variance $\h{\Sigma}_\bb{p}$,
which inherits noise from the simulations. The noise in the covariance
estimator (Eq.~\ref{covest}) and in its inverse $\bbh{\Psi}$ propagate
all the way to the posterior (Eq.~\ref{posteriorbayes}), the parameter
estimate (Eq.~\ref{estimatormean}) and its variance
(Eq.~\ref{estimatorcovariance}). Following
refs.~\citep{DodelsonSchneider13,Taylor12,MasumotoWishart} we compute
the expectation value of Eq.~(\ref{estimatorcovariance}) over
simulations, $\langle\h{\Sigma}_\bb{p}\rangle$, up to $O(1/N_r^2)$ by
expanding Eq.~(\ref{estimatorcovariance}) to quartic order in the
statistical fluctuations of $\bbh{\Psi}$. Denoting the noiseless
parameter covariance, i.e. the usual Fisher matrix, as
$\Sigma_\bb{p}=\bb{F}^{-1}$, we find the result~\footnote{The details
  of the calculation are given in Appendix A.}:

\begin{widetext}
\begin{equation}
\label{quarticdegradation}
\langle\h{\Sigma}_\bb{p}\rangle = \Sigma_\bb{p}\left[1+\frac{N_b-N_p}{N_r}+\frac{(N_b-N_p)(N_b-N_p+2)}{N_r^2}\right] + O\left(\frac{1}{N_r^3}\right).
\end{equation}
\end{widetext}
%
Next, we restrict ourselves to the large $N_r$ limit, and we further
investigate the behavior of the $O(1/N_r)$ term. We consider three
cases \footnote{The details of these calculations are given in Appendix B}:

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/curving_nb.eps}
\caption{Expectation value of the variance of $w$ computed from
  equation (\ref{estimatorcovariance}), shown as a function of
  $1/N_r$. The different symbols and colors correspond to the features
  listed in Table~\ref{featuretable}.  The dashed and solid curves
  show the analytic predictions from equation
  (\ref{quarticdegradation}) at orders $O(1/N_r)$ and $O(1/N_r^2)$,
  respectively. The asymptotic variance $\sigma^2_{w,\infty}$ has been
  computed from a linear regression of $\langle\h{\sigma}^2_w\rangle$
  vs $1/N_r$ for $N_r>500$. The figure clearly shows that terms beyond
  $O(1/N_r)$ need to be considered, unless $N_r\gg N_b$.}
\label{curvingnb}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/scaling_ns.eps}
\caption{The variance of $w$ in the limit of $N_r\rightarrow\infty$
  (measured from the intercept of the fit $\sigma^2_w$ vs $1/N_r$),
  varying the number of simulations $N_s$ used in the ensemble to
  estimate the covariance (eq.~\ref{covest}). We show the dependence
  of $\sigma_{w,\infty}^2(N_s)$ in units of the mean over the 16
  ensembles with different $N_s$, for the power spectrum
  logarithmically binned (black, $N_b=15,\ell\in[100,6000]$), the
  power spectrum linearly binned (red,$N_b=39,\ell\in[100,6000]$) and
  the peak counts (green,$N_b=45,\kappa_0\in[-0.06,0.45]$).  No trend
  with $N_s$ is seen for $N_s\geq 2$, and the differences are only of
  order 1\%.}
\label{wvar_ns}
\end{figure}

@@

Interestingly, we need as few as $N_s=2$ simulations to recover the
right PDF for the small--scale power spectrum.  Figure~\ref{means_ns}
shows that multiple independent $N$--body simulations $N_s\geq2$ are
indeed necessary for measuring the means of feature ensembles to an
accuracy corresponding to 10\% of the statistical error. The number of
required simulations $N_s$ depends on the feature type and ranges from
a few ($N_s=1$ or 2) for the power spectrum at low multipoles
($\ell\lesssim 500$) to $N_s\approx 30-50$ for the power spectrum at
larger multipoles ($\ell\gtrsim1000$) or peak counts above a high
threshold ($\kappa_0\approx0.3$).  On the other hand, relaxing the
required accuracy to 50\% of the statistical error, we find $N_s=2$ to
be always sufficient.

>>>>

\begin{equation}
\label{covest}
\bbh{C} = \frac{1}{N_r-1}\sum_{r=1}^{N_r} (\bbh{d}_r - \bar{\bb{d}}) (\bbh{d}_r - \bar{\bb{d}})^T.
\end{equation}
%
Assuming a normal feature likelihood, together with a flat prior on
the parameter space, the parameter posterior distribution
$\mathcal{L}(\bb{p}\vert\bbh{d}_{\rm obs})$ given an observed instance of
$\bbh{d}$, which we call $\bbh{d}_{\rm obs}$, follows from Bayes' theorem,
%
\begin{equation}
\label{posteriorbayes}
-2\log\mathcal{L}(\bb{p}\vert\bbh{d}_{\rm obs}) = [\bbh{d}_{\rm obs}-\bb{d}(\bb{p})]^T\bbh{C}^{-1}[\bbh{d}_{\rm obs}-\bb{d}(\bb{p})].
\end{equation}
%
For the sake of simplicity, we approximate the posterior as a Gaussian
around its maximum. This corresponds to Taylor-expanding the simulated
feature to first order around a point $\bb{p}_0$ (ideally the maximum
of Eq.~\ref{posteriorbayes}):
%
\begin{equation}
\bb{d}(\bb{p}) \approx \bb{d}_0 + \bb{d}_0^\prime(\bb{p}-\bb{p}_0).
\end{equation}
%
\textbf{\textcolor{blue}{We chose $\bb{p}_0$ to be the triplet $(\Omega_m,w,\sigma_8)=(0.26,-1,0.8)$.}}
To measure the derivatives of the features $\bb{d}'_0$ with respect to
the cosmological parameters, we make use of the public code
\texttt{Nicaea} \citep{Nicaea} for the power spectrum, and we use an
independent simulation set (containing simulations with a variety of
different combinations of $(\Omega_m,w,\sigma_8)$, see Table
\ref{simtable}) for the peak counts.

@@

\begin{equation}
\label{estimatorfisher}
\bbh{F} = \bb{d}_0'^T\bbh{\Psi}\bb{d}_0'.
\end{equation}
%
Here we introduced the familiar Fisher matrix estimator $\bbh{F} =
\bb{d}_0'^T\bbh{\Psi}\bb{d}_0'$ and, for simplicity, we assumed
$\langle\bbh{d}_{\rm obs}\rangle_O=\bb{d}_0$, so that
$\langle(\bbh{d}_{\rm obs}-\bb{d}_0)(\bbh{d}_{\rm obs}-\bb{d}_0)^T\rangle_O=\bb{C}$.  When we perform an observation
$\bbh{d}_{\rm obs}$, the parameter estimate $\bbh{p}$ is a random draw
from a probability distribution with variance $\h{\Sigma}_\bb{p}$,
which inherits noise from the simulations. The noise in the covariance
estimator (Eq.~\ref{covest}) and in its inverse $\bbh{\Psi}$ propagate
all the way to the posterior (Eq.~\ref{posteriorbayes}), the parameter
estimate (Eq.~\ref{estimatormean}) and its variance
(Eq.~\ref{estimatorcovariance}). Following
refs.~\citep{DodelsonSchneider13,Taylor12,MasumotoWishart} we compute
the expectation value of Eq.~(\ref{estimatorcovariance}) over
simulations, $\langle\h{\Sigma}_\bb{p}\rangle$, up to $O(1/N_r^2)$ by
expanding Eq.~(\ref{estimatorcovariance}) to quartic order in the
statistical fluctuations of $\bbh{\Psi}$. Denoting the \textbf{\textcolor{blue}{true}}
parameter covariance, i.e. the usual \textbf{\textcolor{blue}{inverse}} Fisher matrix, as
$\Sigma_\bb{p}=\bb{F}^{-1}$, we find the result~\footnote{The details
  of the calculation are given in Appendix A.}:

\begin{widetext}
\begin{equation}
\label{quarticdegradation}
\langle\h{\Sigma}_\bb{p}\rangle = \Sigma_\bb{p}\left[1+\frac{N_b-N_p}{N_r}+\frac{(N_b-N_p)(N_b-N_p+2)}{N_r^2}\right] + O\left(\frac{1}{N_r^3}\right).
\end{equation}
\end{widetext}
%
\textbf{\textcolor{blue}{Although we truncated the expansion to second order in $1/N_r$, an exact expression for $\langle\h{\Sigma}_\bb{p}\rangle$ has been proposed by \citep{Taylor14}}}
\begin{equation}
\label{empiricalexact}
\textcolor{blue}{\langle\h{\Sigma}_\bb{p}\rangle_{\rm empirical} = \Sigma_\bb{p}\left(\frac{N_r-2}{N_r-N_b+N_p-2}\right)}
\end{equation}
%
\textbf{\textcolor{blue}{This empirical expression reduces to equation (\ref{quarticdegradation}) when expanded at order $O(1/N_r^2)$ but, to our knowledge, no first principles proof of its correctness exists.}} 
Next, we restrict ourselves to the large $N_r$ limit, and we further
investigate the behavior of the $O(1/N_r)$ term. We consider three
cases \footnote{The details of these calculations are given in Appendix B}:

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/curving_nb.eps}
\caption{Expectation value of the variance of $w$ computed from
  equation (\ref{estimatorcovariance}), shown as a function of
  $1/N_r$. The different symbols and colors correspond to the features
  listed in Table~\ref{featuretable}.  The dashed and this solid curves
  show the analytic predictions from equation
  (\ref{quarticdegradation}) at orders $O(1/N_r)$ and $O(1/N_r^2)$,
  respectively. \textbf{\textcolor{blue}{The thick solid curves show the empirical predictions from equation (\ref{empiricalexact})}}. 
  The asymptotic variance $\sigma^2_{w,\infty}$ has been
  computed from a linear regression of $\langle\h{\sigma}^2_w\rangle$
  vs $1/N_r$ for $N_r>500$. The figure clearly shows that terms beyond
  $O(1/N_r)$ need to be considered, unless $N_r\gg N_b$.}
\label{curvingnb}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/scaling_ns.eps}
\caption{The variance of $w$ in the limit of $N_r\rightarrow\infty$
  (measured from the intercept of the fit $\sigma^2_w$ vs $1/N_r$),
  varying the number of simulations $N_s$ used in the ensemble to
  estimate the covariance (eq.~\ref{covest}). We show the dependence
  of $\sigma_{w,\infty}^2(N_s)$ in units of the mean over the \textbf{\textcolor{blue}{union of}} 16
  ensembles with different $N_s$ \textbf{\textcolor{blue}{and $N_r=1024$}}, for the power spectrum
  logarithmically binned (black, $N_b=15,\ell\in[100,6000]$), the
  power spectrum linearly binned (red,$N_b=39,\ell\in[100,6000]$) and
  the peak counts (green,$N_b=45,\kappa_0\in[-0.06,0.45]$).  No trend
  with $N_s$ is seen for $N_s\geq 2$, and the differences are only of
  order 1\%.}
\label{wvar_ns}
\end{figure}

@@

Interestingly, we need as few as $N_s=2$ simulations to recover the
right PDF for the small--scale power spectrum.  Figure~\ref{means_ns}
shows that multiple independent $N$--body simulations $N_s\geq2$ are
indeed necessary for measuring the means of feature ensembles to an
accuracy corresponding to 10\% of the statistical error. The number of
required simulations $N_s$ depends on the feature type and ranges from
a few ($N_s=1$ or 2) for the power spectrum at low multipoles
($\ell\lesssim 500$) to $N_s\approx 30-50$ for the power spectrum at
larger multipoles ($\ell\gtrsim1000$) or peak counts above a high
threshold ($\kappa_0\approx0.3$).  On the other hand, relaxing the
required accuracy to 50\% of the statistical error, we find $N_s=2$ to
be always sufficient. \textbf{\textcolor{blue}{As pointed out by \citep{NbodySample}, the box size used for the $N$--body simulations can also play an important role in the accuracy of the power spectrum ensemble means.}}
